---
---

::: {.callout-tip}
## Learning Goals {.unnumbered}

By the end of this tutorial, you should be able to:

- Understand key factors related to employee attrition and their potential impact on the business.
- Apply clustering techniques (K-means, PCA) to uncover segments of employees with different characteristics.
- Interpret and compare the resulting clusters in terms of job satisfaction, income, tenure, and attrition risk.
- Evaluate how well the clustering reflects meaningful differences using visualizations and summary statistics.
:::

## The Business Challenge

### The Topic: Why are employees leaving — and how can we group them to better understand attrition? {.unnumbered}

Every time an employee leaves, companies incur real costs — from lost productivity to recruitment and training. Understanding **which types of employees are most at risk of leaving** is a key priority for HR analytics.

In this tutorial, we’ll use clustering techniques to group employees based on their characteristics and see whether some groups are more likely to quit. This lets us move beyond individual-level predictors to identify broader patterns and profiles — a common strategy in **descriptive analytics**.

We'll also introduce **Principal Component Analysis (PCA)** — a method for reducing complexity while keeping the structure of the data intact.

### The Data: HR Analytics Data on 1,470 Employees {.unnumbered}

We'll be working with an anonymized dataset from a fictional company’s HR department. It includes detailed information on 1,470 employees, including:

- **Personal characteristics**: `age`, `gender`, `marital_status`, `distance_from_home`
- **Job-related features**: `job_satisfaction`, `monthly_income`, `years_at_company`, `department`
- **Outcome**: `attrition` – whether the employee has left the company

This gives us a rich set of variables to explore patterns in attrition and test whether employee segments differ in meaningful ways.

### The Method: Clustering and PCA for Descriptive Segmentation {.unnumbered}

Our goal is to **segment employees** into groups with similar profiles — and examine whether some segments are more prone to attrition. To do this, we’ll:

1. **Prepare and normalize** the data so that all variables are on comparable scales  
2. Use **K-means clustering** to group employees based on features like satisfaction, income, and commute distance  
3. Use **PCA** to reduce the number of variables while preserving the main structure in the data  
4. Interpret the resulting clusters — comparing their average characteristics and attrition rates

We'll also explore how to choose the **right number of clusters**.

### Loading the R packages and data

Here's the packages we will need to complete the exercises:

```{r}
library(tidyverse)  # data manipulation and plotting
library(janitor)    # cleaning column names
library(recipes)    # For data preprocessing
library(cluster)    # clustering
library(factoextra) # k-means visualization and cluster evaluation
library(broom)      # tidy output of clustering results
```

Let's load the data that we will use today:

```{r}
df <-
    read_csv("data/hr_attrition.csv") |> 
    # tidy up column names to snake_case 
    clean_names()
```

## Prepare these Exercises before Class 

Prepare these exercises before coming to class. Plan to spend 30 minutes on these exercises.

### Exercise 1: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

### Exercise 2: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

### Exercise 3: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

### Exercise 4: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

### Exercise 5: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

## In-Class Exercises

You will discuss these exercises in class with your peers in small groups and with your tutor. These exercises build from the exercises you have prepared above, you will get the most value from the class if you have completed those above before coming to class.

### Exercise 1: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

### Exercise 7: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

### Exercise 8: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

### Exercise 9: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

### Exercise 10: Title

This is the text for the question, with its accompanying code 

```{r}
# Example code for the question
1 + 1
```

```{r}
# Write your answer here
```

