<!-- question-type: inclass -->
### [Advanced, Optional] Exercise 7: Segmenting with All Variables

::: {.callout-warning collapse=true}
## Only complete if > 40 mins left in the tutorial!

This exercise takes a while to complete. If there is less than 40 mins to go in class, skip this exercise and proceed to Exercise 8.
:::

::: {.callout-warning collapse=true}
## What comes next is tough!

In the exercises that follow, you’ll walk through how to do K-means clustering using **both numeric and categorical features**, with **dimensionality reduction via Principal Component Analysis (PCA)**.

This is advanced content — we don’t expect you to write this code on your own nor will a detailed example such as this be expected in an assessment situation.
We want to show you what **is** possible building off the skills you have acquired above.
Follow along with your tutor and focus on building intuition for:

- What each step does
- Why we need it
- How it connects to what you’ve learned so far

Exercises after the data prepation will feel very familiar — only the somewhat involved data preparation is new.
:::

So far, we've limited our clustering to just five numeric features.  
But in reality, companies might want to use **all available information** — including categorical variables like job role, marital status, and department — to uncover richer segmentation patterns.

That introduces two new challenges:

1. **K-means clustering only works with numeric inputs.**  
   So how do we use categorical variables?

2. **K-means works best with fewer dimensions.**  
   But after we include all those dummy variables, we get many more!

This exercise guides you through a workflow to overcome both challenges.



**(a)** Run the data preparation pipeline below that:

- Removes `employee_number` (a unique ID)
- Removes `attrition` (our outcome variable — we want to use it *after* clustering)
- Converts categorical variables into dummy (one-hot encoded) columns
- Normalises all numeric columns

```{r}
#| eval: false
rec <- 
    recipe(~ ., data = df) |>
    # Remove these variables from processing pipeline
    update_role(employee_number, new_role = "id") |>
    update_role(attrition, new_role = "outcome") |>
    # Drop variables that are constant across all rows
    step_zv(all_predictors()) |>
    # Create "Dummmy Variables" from Categoricals
    step_dummy(all_nominal_predictors()) |>
    # normalize all numerics, as before, but across all
    step_normalize(all_numeric_predictors())

df_pca_all <- 
    prep(rec) |> 
    bake(new_data = NULL) |>
    # drop these from the PCA
    select(-employee_number, -attrition)

```

**(b)** What does the `step_dummy()` line do?
Why does this solve the problem we mentioned above — that K-means requires numeric inputs?

**(c)** The pipeline also includes `step_normalize(all_numeric_predictors())`, but this doesn’t affect the dummy variables.
Why might that be a good thing? What could go wrong if we normalised dummy variables?

**(d)** Why do we remove `employee_number` and `attrition` after the `bake()` step?

**(e)** Run PCA on the dataframe created in (a) and investigate the output. Think about how many components you would want to keep for the K-means analysis.

```{r}
#| eval: false
pca_model <- 
    prcomp(df_pca_all, 
           center = FALSE, 
           scale. = FALSE
           )
summary(pca_model)

fviz_contrib(pca_model, choice = "var", axes = 2)


```

**(f)** Run the code below to keep the first 2 Principal Components. Can you explain why we chose to keep only the first two?

```{r}
#| eval: false 
df_pca_2 <- as_tibble(pca_model$x[, 1:2])
```

**(g)** Use the Elbow method to determine the number of clusters you want to use in your K-means analysis. How many clusters would you choose?

```{r}
#| eval: false
fviz_nbclust(
  df_pca_2, 
  kmeans,
  method = "wss",     # Within-cluster sum of squares
  k.max = 10,
  nstart = 10
)
```

**(h)** Run the code below to run K-means clustering with `k=3` clusters.

```{r}
#| eval: false
set.seed(123)
kmeans_pca <- kmeans(df_pca_2, centers = 3, nstart = 10)
```

**(i)** Visualise the PCA clustering results using the code below. Did the three cluster model perform better compared to Exercises 4 and 6? Try and explain why.

```{r}
#| eval: false
fviz_cluster(kmeans_pca, data = df_pca_2)
```

<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Understand the challenges of clustering with both numeric and categorical data.
- Recognize how dummy coding and scaling prepare data for clustering.
- Understand the role of PCA in reducing dimensionality when working with high-dimensional data.
- Be exposed to real-world workflows combining preprocessing, PCA, and K-means clustering.

✅   **Core Concepts to Highlight**

- Dummy coding (`step_dummy()`): Converts categorical variables into numeric format so they can be used in K-means.
- Why not normalize dummies?: 0/1 coding should be preserved to retain their meaning and avoid distorting distance.
- Dimensionality and PCA: High-dimensional clustering can be unstable; PCA helps compress relevant variation into fewer features.
- Roles in recipes: Using `update_role()` to exclude outcome/ID variables from preprocessing.
- Pipeline logic: Walk through the full transformation > PCA > clustering chain.


💬 **Suggested In-Class Prompts** (if needed)

"What would happen if we left the categorical variables as-is?"

"Why might a high number of dimensions be a problem for clustering?"

"What does the PCA plot tell us about the data’s structure?"

📌 **Common Misunderstandings**

"Dummy variables are numeric, so they should be scaled."	

* Dummy variables are already on a comparable 0/1 scale. Scaling distorts their meaning.

"PCA makes the clustering more accurate."	

* PCA simplifies the space; it may improve interpretability or speed, but not always clustering accuracy.

"We can use attrition in the clustering."	

* No — this would make the clustering predictive rather than exploratory. We use it after clustering to describe the segments.

"All data prep is done automatically."	

* Emphasize the explicit choices in the recipe: what’s included, excluded, and how each transformation works.

"Two components = 2 variables."	

* PCA components are linear combinations of many variables, not a one-to-one replacement.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a)** Data Prep Pipeline

The provided code correctly does the following:

- Uses `recipes::recipe()` to set up a preprocessing workflow
- Removes `employee_number` (set as "id" role)
- Removes `attrition` (set as "outcome" role)
- Handles constant columns using `step_zv()`
- Encodes all categorical variables via `step_dummy()`
- Normalizes all numeric variables (excluding dummy variables) via `step_normalize()`
- Applies `prep()` and `bake()` to generate the transformed dataset

**(b)** What does `step_dummy()` do and why is it helpful?

`step_dummy()` transforms each categorical variable into one or more binary columns (0/1) — one for each category level (except the reference category, unless one_hot = TRUE).

This solves the problem that K-means only works with numeric inputs.
By converting categoricals to numeric 0/1 dummies, we can now use them in the clustering algorithm.

**(c)** Why not normalize dummy variables?

Dummy variables only take values of 0 or 1. If we normalize them:

- The 0/1 scale becomes continuous (e.g., -0.5 to +0.5), which changes the meaning.
- It could distort distance calculations in K-means by over-emphasizing categorical differences.

For example, if a rare category becomes a high-magnitude value after normalization, it may exaggerate its influence in clustering.

Thus, dummy variables are best left as 0/1.

**(d)** Why do we remove employee_number and attrition after `bake()`?

These variables were excluded from the preprocessing steps via role assignment — but they still remain in the baked dataset.

We remove them after `bake()` because:

- `employee_number` is a unique ID and doesn't help clustering (and could distort distance).
- `attrition` is an outcome variable, not a predictor — we use it after clustering to describe the segments, not to define them.

**(e)** PCA Output

The PCA model reveals how much variance is explained by each principal component.

Key observations:

- The first few components (e.g., PC1 and PC2) typically capture a large share of the variation.
- Component contribution plots help interpret what kinds of variables dominate each component.

**(f)** Why keep only 2 components?

Choosing only 2 components as these pick up the most important features from back in Ex 1.

**(g)** Choosing k via the Elbow method

The Elbow plot helps determine the optimal number of clusters by showing diminishing returns in reducing within-cluster sum of squares.

Look for a point where the curve "bends" — this is often between k = 3 to 4.

Here, k = 3 is where there is a clear bend.

**(h)** Run K-means clustering

Code is correct. Students should observe that:

- The `kmeans()` function uses the PCA-transformed data
- `set.seed(123)` ensures reproducibility
- `nstart = 10` gives multiple initializations

**(i)** Cluster Visualisation

The PCA cluster plot should show:

- Clearer separation between clusters than in previous exercises
- More balanced or tighter groupings, depending on how well PCA compressed the data

The performance may feel better due to using more features and reducing noise through PCA.
However, interpretation of clusters becomes harder, since we're no longer in the original feature space.

:::

:::

:::
<!-- END PROFILE:r-solutions -->