<!-- question-type: inclass -->
### Exercise 5: How many clusters should we use?

Clustering requires us to pick a value for `k` — the number of clusters. But how do we decide what’s best?

There are several methods we can use. Each has a different way of deciding whether the clusters are "tight" and "well separated."

**(a)** What do you think could go wrong if we choose too few or too many clusters?

**(b)** One common approach we encountered in the lecture is the **elbow method**.  
What is the elbow method trying to do, in plain language?

**(c)** Run the code below and look at the plot.  
Where do you see a clear "elbow" or bend in the curve?  
Based on the plot, what value of `k` would you choose?

```{r}
#| eval: false
fviz_nbclust(
  df_scaled, 
  kmeans,
  method = "wss",     # Within-cluster sum of squares
  k.max = 10,
  nstart = 10
)
```

**(d)** One alternative to the elbow method is the silhouette method which looks at how well each point fits within its cluster vs. other clusters.
Run the code below and report which value of `k` this method prefers.

```{r}
#| eval: false
fviz_nbclust(
  df_scaled, 
  kmeans,
  method = "silhouette"
)
```

**(e)** Another alternative to the elbow method is the gap statistic which compares your clustering to what you’d expect from random data.
Run the code below and report which value of `k` this method prefers.

```{r}
#| eval: false
fviz_nbclust(
  df_scaled,
  kmeans,
  method = "gap_stat",
  nstart = 25,
  iter.max = 50
)
```

**(f)** Do all three methods agree on the best value of k? If not, how would you decide?

**(g)** Based on your answer, pick a value for k, rerun the kmeans() function, and briefly describe the new clustering results.

```{r}
#| eval: false
set.seed(123)
kmeans_model <- kmeans(df_scaled, centers = YOUR_CHOICE, nstart = 10)

fviz_cluster(
  kmeans_model,
  data = df_scaled,
  palette = "jco",
  ggtheme = theme_minimal()
)
```


<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Understand the importance of selecting an appropriate number of clusters in K-means.
- Explain in plain language how common methods for choosing k (elbow, silhouette, gap statistic) work.
- Apply each method in R and interpret the outputs.
- Make a justified choice of k based on multiple criteria.

✅   **Core Concepts to Highlight**

- K-means clustering requires a user-specified number of clusters (k).
- Within-Cluster Sum of Squares (WSS) as a measure of compactness.
- Elbow method: diminishing returns in WSS reduction.
- Silhouette method: cohesion vs. separation.
- Gap statistic: comparison to a reference distribution.
- Model selection trade-offs: interpretability, stability, business relevance.


💬 **Suggested In-Class Prompts** (if needed)

“Can someone explain what the elbow method is trying to find, without using the word ‘elbow’?”

“Why might different methods recommend different values for k? What does that tell us?”

“Suppose silhouette says 2 but elbow and gap say 3 — what would you choose, and why?”

“Does your chosen number of clusters make sense in business terms?”

📌 **Common Misunderstandings**

"Always choose the number with lowest WSS" – forgetting that WSS always drops with higher k.

Thinking the goal is to make all clusters the same size – remind them that’s not required.

Treating method output as absolute truth – emphasize judgment and context.

Not rerunning kmeans() after choosing k – they need to redo clustering using their chosen value.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

### (a) What could go wrong with too few or too many clusters?

- **Too few clusters**: We may group together employees who are actually quite different. This hides useful variation and makes it hard to take targeted action.
- **Too many clusters**: We might split up natural groupings and end up with tiny, unstable segments that are hard to interpret or act on.


**(b)** What is the elbow method doing, in plain language?

- The elbow method looks at how much **within-cluster variation** (WSS) drops as we increase `k`.
- As we increase `k`, WSS will naturally decrease — but after a point, the **improvement slows down**.
- The "elbow" is the value of `k` where the rate of improvement flattens — indicating that adding more clusters doesn’t improve fit by much.


**(c)** What value of `k` would you choose based on the elbow plot?

```{r}
#| eval: true

# remark: scaled data is stored in data, you can load it as
# df_scaled <- 
#   read_csv("data/hr_attrition_scaled.csv")

fviz_nbclust(
  df_scaled, 
  kmeans,
  method = "wss",     # Within-cluster sum of squares
  k.max = 10,
  nstart = 10
)
```

Students might give slightly different answers, but **common responses** include:

- If there’s a decrease in steepening around `k = 2` or `k = 3`, this is often a good choice.
- Encourage students to **justify** their answer

**(d)**

```{r}
#| eval: true
fviz_nbclust(
  df_scaled, 
  kmeans,
  method = "silhouette"
)
```

2 clusters.

- Students should read this directly from the **peak** of the silhouette plot.
- Remind them that **higher silhouette values** mean **tighter and more distinct** clusters.

**(e)**

```{r}
#| eval: false
fviz_nbclust(
  df_scaled,
  kmeans,
  method = "gap_stat",
  # nstart of 25 stops the warning methods appearing for no gains
  nstart = 25,
  iter.max = 50
)
```

- The **gap statistic** chooses the `k` where the gap between real and random data is largest.
- This is  shown with **a bar or point marker** on the plot.
- Common answer: in this example 1.

**(f)**

- Elbow suggest **k = 2 or 3**
- Silhouette prefers **k =  2**
- Gap statistic suggest **k = 1**
- Students should reflect critically. For instance:
  - "If both silhouette and gap agree on 3, but elbow suggests 1, I might choose 2."
  - "If k = 3 produces more interpretable clusters, that's also a good reason to choose it."

**(g)**

```{r}
set.seed(123)
kmeans_model <- kmeans(df_scaled, centers = 2, nstart = 10)

fviz_cluster(
  kmeans_model,
  data = df_scaled,
  palette = "jco",
  ggtheme = theme_minimal()
)

# if you have time, add these clusters back to the data and look at what they identify (tho one can infer this from Q4)
```

:::

:::

:::
<!-- END PROFILE:r-solutions -->