<!-- question-type: prepare -->
### Exercise 2: Identifying Some Preliminary Patterns

::: {.callout-warning collapse=true}
### Why We're Not Giving You Template Code

You've had plenty of practice with `dplyr` tools like `group_by()`, `summarise()`, and `mutate()`. This question is designed to help you build fluency in writing tidyverse code from scratch.

It's normal to feel unsure at first—but struggling a little helps solidify your skills. If you're stuck, re-use patterns from previous tutorials.
:::


Let's start exploring the dataset to identify patterns in attrition. 
Our goal is to understand where attrition is most common and what kinds of employees are more likely to leave.

**(a)** What is the overall attrition rate in the dataset?

**(b)** Which departments or job roles seem to have the highest attrition? Use proportions, not just counts.

**(c)** How do the following characteristics differ between employees who left and those who stayed?
Use averages to compare:

- Job satisfaction
- Monthly income
- Years at company
- Distance from home
- Age

**(d)** Based on your findings, describe the typical profile of an employee who leaves. Are there any surprises?


<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Apply grouped summarisation to examine attrition patterns across categories.
- Use pivot_wider() to restructure data for clearer comparison.
- Interpret proportions and identify roles with high attrition rates.

✅   **Core Concepts to Highlight**

- Relative vs absolute comparisons: Why proportions (shares) are often more insightful than raw counts.
- Data reshaping: Using `pivot_wider()` to improve readability and enable side-by-side comparisons.
- Sorting by relevance: How `arrange()` can surface the most important categories (e.g. high attrition)


💬 **Suggested In-Class Prompts** (if needed)

“Which job roles have the highest proportion of attrition?”

“Why might it be misleading to just use counts (n) instead of proportions?”

“If we wanted to visualise this, what type of chart would help?”

“Why is it important to compare within group (e.g., share of leavers among each role) rather than across the whole company?”

📌 **Common Misunderstandings**

- Sorting by `n` instead of `share`: Students often default to sorting by total count, missing the within-group interpretation.
- Confusing base rates: Some students assume high total attrition count = high attrition rate, without considering role size.
- Zero values: Forgetting `values_fill = 0` may lead to NA values which confuse students when interpreting results.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a)**

```{r}
df |> 
  group_by(department, attrition) |> 
  summarise(n = n(), .groups = "drop") |> 
  mutate(share = n / sum(n))
```


**(b)**

```{r}
df |> 
  group_by(job_role, attrition) |> 
  summarise(n = n(), .groups = "drop") |> 
  mutate(share = n / sum(n)) |> 
  pivot_wider(
    names_from = attrition,
    values_from = share,
    values_fill = 0
  ) |> 
    arrange(desc(Yes))
```


**(c)**

```{r}
#REMARK: across with an anonymous function is new. its faster to write
# feel free to do it the manual way if more comfortable
df |> 
  group_by(attrition) |> 
  summarise(
    across(
      c(job_satisfaction, monthly_income, years_at_company, distance_from_home, age),
      \(x) mean(x, na.rm = TRUE)
    ),
    .groups = "drop"
  )
```


**(d)**

Employees who leave the company tend to:

- Have lower job satisfaction
- Be paid less
- Have fewer years at the company
- Live further from the office
- Be slightly younger
- Be in Sales or more junior Research roles

:::

:::

:::
<!-- END PROFILE:r-solutions -->