<!-- question-type: inclass -->
### Exercise 6: Segmentation with Dimensionality Reduction

We now explore whether simplifying our dataset — by reducing the number of dimensions — might help improve clustering.

Right now, we're clustering using all five numeric features. But some of these might contain similar information. 
Dimensionality reduction helps us find a smaller number of new features (called components) that capture most of the variation in the data.
Once we have reduced dimensionality, we can then run Segmentation on this reduced set of features.
This will prove helpful in the next question when we want to use **all** variables in the dataset

One common method of dimensionality reduction is Principal Component Analysis (PCA).


**(a)** Ask your tutor to briefly explain the intuition behind the PCA method and how it works.
Then, in your own words, write down one thing that PCA is trying to do.

**(b)** Run the code below to apply PCA to the `df_scaled` data.
Then use summary(pca_model) to inspect how much variance is explained by the first few components.

```{r}
#| eval: false
pca_model <- 
    prcomp(df_scaled,
           # because we have already done each of these
           # set them to false
           center = FALSE,   # if TRUE sets means to 0
           scale. = FALSE)   # IF TRUE sets std dev to 1
summary(pca_model)
```

**(c)** After looking at the output, suppose a colleague suggests "use 3 principal components". Explain why he might make that suggestion.

**(d)** Another way to choose how many components to keep is to look at which variables are being absorbed into each components.
Run the code below, and then decide how many components you want to keep, justifying your decision.

```{r}
#| eval: false
# PC1 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 1)

# PC2 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 2)

# PC3 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 3)

# PC4 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 4)

# PC5 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 5)
```

**(e)** Create a new PCA-transformed dataset with only the components you selected above by completing the code below:

```{r}
#| eval: false
df_pca <- as_tibble(pca_model$x[, 1:YOUR_CHOICE])
```

**(f)** Run K-means on `df_pca` using 3 clusters.

```{r}
#| eval: false
kmeans_pca <- kmeans(df_pca, centers = 3, nstart = 10)
```

**(g)** Visualise the PCA clustering results using the code below. Did the three cluster model perform better compared to Exercise 4? Try and explain why.

```{r}
#| eval: false
fviz_cluster(kmeans_pca, data = df_pca)
```


<!-- BEGIN PROFILE:r-teaching-guide -->
::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective** 
Students should:

- Understand the intuition behind dimensionality reduction using PCA.
- Identify when and why PCA can improve clustering.
- Practice applying PCA to scaled numeric data.
- Use component loadings and variance explained to decide how many components to keep.
- Evaluate the effect of dimensionality reduction on clustering quality.

✅   **Core Concepts to Highlight**

- Dimensionality reduction: Simplifying a dataset by summarizing information across correlated variables.
- Principal Component Analysis (PCA): A method that transforms original variables into a smaller set of orthogonal components that explain the most variance.
- Variance explained: Interpreting PCA output to understand how much of the original data structure is retained.
- Component loadings: Using variable contributions to understand what each PC represents.
- Segmentation in reduced space: Running k-means on the transformed data and comparing clustering results.


💬 **Suggested In-Class Prompts** (if needed)

“What kinds of situations call for dimensionality reduction?”

“Can you think of examples where two variables might be telling us the same thing?”

“What are the tradeoffs between keeping more or fewer principal components?”

“Does clustering look more stable or interpretable after PCA?”

📌 **Common Misunderstandings**

- Misinterpreting PCA axes: Students may think PCs are just the original variables reordered. Emphasize that each PC is a weighted combination of the original variables.
- Ignoring the need to scale data: PCA assumes data is centered and scaled; remind students we did this in a previous exercise.
- Using too many components: Keeping all components defeats the purpose of dimensionality reduction — we want a simplified representation.
- Comparing cluster plots directly: Students may assume PCA space should "look" like original data space. Encourage focus on separation, not interpretability of axes.

:::

:::

:::
<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->
::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a)**
PCA is a technique for dimensionality reduction. It creates new variables (called *principal components*) that are combinations of the original variables. These components capture the largest possible variance in the data, with each new component orthogonal (independent) from the previous ones.

PCA is trying to:  
- Reduce the number of variables while retaining as much information as possible  
- Identify patterns in the data by summarising correlated variables

**(b)** 

```{r}
#| eval: true
pca_model <- prcomp(df_scaled, center = FALSE, scale. = FALSE)
summary(pca_model)
```
This shows how much variance each principal component explains.

- PC explain ~37% of the variance
- PC2 ~20%
- PC3 ~20%
- PC4 and PC5 less

This suggests the first 3 components explain around 80% of the total variance.

**(c)** Why choose 3 components?

Because the first 3 components together explain most of the variance in the data (usually >75–80%), they provide a good summary while reducing noise and complexity. Including more components adds little additional insight and may reintroduce noise.

**(d)** Variable contributions to each component

```{r}
#| eval: true
# PC1 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 1)

# PC2 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 2)

# PC3 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 3)

# PC4 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 4)

# PC5 variable contributions
fviz_contrib(pca_model, choice = "var", axes = 5)
```

From these plots, we can see:

PC1 may be driven mostly by `monthly_income`, `years_at_company` and `age`

PC2 by `distance_from_home` and `job_satisfaction`
 
PC3 again by by by `distance_from_home` and `job_satisfaction`

Since the first two components clearly loads on a different set of variables, this supports the case for keeping 2 components.

**(e)** 

```{r}
#| eval: true
df_pca <- as_tibble(pca_model$x[, 1:2])
```

**(f)** 

```{r}
#| eval: true
set.seed(123)
kmeans_pca <- kmeans(df_pca, centers = 3, nstart = 10)
```

**(g)**

```{r}
#| eval: true
fviz_cluster(kmeans_pca, data = df_pca)
```


We find:

The clusters are more clearly separated in PCA space

There's less overlap than in the original 5D space

Interpretation:
PCA helps reduce noise and overlapping signals, so the clusters become tighter and better defined. Especially when some variables were correlated, PCA unbundles that structure and makes segmentation easier.

:::

:::

:::
<!-- END PROFILE:r-solutions -->